---
title: "Installation Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installation Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

Welcome to **leo.sc**! This guide outlines the recommended steps to install the package and its dependencies. Since `leo.sc` relies on various CRAN, Bioconductor, and GitHub-hosted packages, a structured installation approach ensures a smooth setup.

## Prerequisites

Ensure you have a recent version of **R (>= 4.1.0)** installed. We highly recommend using [RStudio](https://posit.co/download/rstudio-desktop/) or [radian](https://github.com/randy3k/radian) running inside [VS Code](https://code.visualstudio.com/) as your integrated development environment (IDE).

You will need the `devtools` or `remotes` package to install from GitHub:

```{r install-devtools, eval=FALSE}
if (!requireNamespace("devtools", quietly = TRUE)) {
  install.packages("devtools")
}
```

## Standard Installation

The most straightforward method to install `leo.sc` is directly from GitHub. The `devtools::install_github` function will automatically attempt to resolve and install dependencies listed in the package configuration.

```{r install-leosc, eval=FALSE}
devtools::install_github("laleoarrow/leo.sc")
```

## Troubleshooting Dependencies

Bioinformatics packages often have complex dependency trees. If the standard installation fails due to missing packages, follow these targeted troubleshooting steps:

### 1. Bioconductor Dependencies

Several core functions in `leo.sc` depend on Bioconductor infrastructure (e.g., `clusterProfiler`, `org.Hs.eg.db`, `enrichplot`). If these fail to install automatically, install them manually using BiocManager:

```{r install-bioc, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

BiocManager::install(c(
  "clusterProfiler",
  "org.Hs.eg.db",
  "enrichplot"
))
```

### 2. GitHub-Hosted Dependencies

`leo.sc` relies on other custom packages, notably `leo.basic` and `DoubletFinder`. While these are declared as remotes, network timeouts or API limits can sometimes interrupt their installation. You can install them manually beforehand:

```{r install-remotes, eval=FALSE}
devtools::install_github("chris-mcginnis-ucsf/DoubletFinder")
devtools::install_github("laleoarrow/leo.sc")
```

## Verification

Once the installation completes, verify that the package loads successfully into your R session:

```{r verify_install, eval=FALSE}
library(leo.sc)
cli::cli_alert_success("leo.sc loaded successfully!")
```

If you encounter persistent issues during installation, please open an issue on the [leo.sc GitHub repository](https://github.com/laleoarrow/leo.sc/issues).
