---
title: "FIP Case Study: MS4A1 (CD20) in PBMC3k"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{FIP Case Study: MS4A1 (CD20) in PBMC3k}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)

find_leo_sc_root <- function(start = getwd()) {
  cur <- normalizePath(start, winslash = "/", mustWork = TRUE)
  repeat {
    desc <- file.path(cur, "DESCRIPTION")
    if (file.exists(desc)) {
      desc_lines <- readLines(desc, warn = FALSE)
      if (any(grepl("^Package:\\s*leo\\.sc\\s*$", desc_lines))) {
        return(cur)
      }
    }
    parent <- dirname(cur)
    if (identical(parent, cur)) {
      return(NULL)
    }
    cur <- parent
  }
}

pkg_root <- find_leo_sc_root()
if (!is.null(pkg_root) && requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else if (!requireNamespace("leo.sc", quietly = TRUE)) {
  stop("Please install leo.sc, or render from a leo.sc source checkout.")
}

silico_ko <- get("silico_ko", envir = asNamespace("leo.sc"), inherits = FALSE)
```

## Context

This case study applies FIP (`silico_ko()`) to the classic **PBMC3k** dataset
([10x Genomics, 2016](https://support.10xgenomics.com/single-cell-gene-expression/datasets/1.1.0/pbmc3k))
to infer transcriptional programs associated with high endogenous expression of **MS4A1 (CD20)**.

MS4A1 encodes a membrane-spanning 4A family protein that is exclusively expressed on **B cells**
and is the pharmacological target of anti-CD20 monoclonal antibodies (e.g., Rituximab). Its
canonical roles include BCR signal transduction, B cell activation, and humoral immunity — making it
an ideal positive-control gene for FIP: enrichment results should cleanly recapitulate known B cell
biology, directly validating the method.

For a minimal, dependency-free demonstration, see the [pbmc_small quick demo](tutorial-sko.html).

## Data

```{r data-setup}
if (!requireNamespace("SeuratData", quietly = TRUE)) {
  stop("SeuratData not found. Install with: devtools::install_github('satijalab/seurat-data')")
}

suppressMessages(SeuratData::InstallData("pbmc3k"))
suppressMessages(data("pbmc3k.final", package = "SeuratData"))

# pbmc3k.final carries pre-computed Seurat clusters and canonical cell-type labels.
# Reference: https://satijalab.org/seurat/articles/pbmc3k_tutorial
srt <- pbmc3k.final
srt$cell_anno <- as.character(srt$seurat_annotations)

cli::cli_alert_info(
  "Dataset: {ncol(srt)} cells x {nrow(srt)} genes | {length(unique(srt$cell_anno))} annotated cell types"
)

ct_tbl <- as.data.frame(sort(table(srt$cell_anno), decreasing = TRUE))
colnames(ct_tbl) <- c("Cell type", "N")
knitr::kable(ct_tbl, caption = "PBMC3k cell-type composition")
```

## FIP: MS4A1 Knock-High Analysis

```{r run-sko}
res <- silico_ko(
  all                   = srt,
  gene                  = "MS4A1",
  sko_mode              = "ko", # compare high- vs low-expressors
  cell_col              = "cell_anno",
  filter_cell_threshold = 10, # exclude cell types with < 10 MS4A1-expressing cells
  pct_threshold         = 0.1, # select top/bottom 10% per cell type
  deg_method            = "default", # Wilcoxon rank-sum via Seurat::FindMarkers
  enrichment_method     = "ORA",
  enrichment_bg         = "GO",
  simplify              = TRUE
)
```

## Results

### Cell-type composition of ***g***<sup>high</sup>

```{r cell-composition}
knitr::kable(
  res$cell_rato,
  caption = "Cell-type composition of MS4A1-high group (g<sup>high</sup>)"
)
```

MS4A1 expression is essentially confined to **B cells** in PBMCs — the ***g***<sup>high</sup> group is therefore
expected to be dominated by B cells, serving as a built-in biological sanity check for FIP.

### Differential expression

```{r deg-summary}
deg <- tibble::rownames_to_column(res$deg_results, "gene")

n_up <- sum(deg$avg_log2FC > 0 & deg$p_val_adj < 0.05, na.rm = TRUE)
n_down <- sum(deg$avg_log2FC < 0 & deg$p_val_adj < 0.05, na.rm = TRUE)
cli::cli_alert_info("DEG summary — up in MS4A1-high: {n_up} | down: {n_down}")

top5 <- function(df, direction = "up", n = 5) {
  df <- if (direction == "up") df[df$avg_log2FC > 0, ] else df[df$avg_log2FC < 0, ]
  df <- df[!is.na(df$p_val_adj), ]
  df <- df[order(abs(df$avg_log2FC), decreasing = TRUE), ]
  head(df[, c("gene", "avg_log2FC", "p_val_adj")], n)
}

knitr::kable(top5(deg, "up"), digits = 3, caption = "Top genes enriched in MS4A1-high cells")
knitr::kable(top5(deg, "down"), digits = 3, caption = "Top genes depleted in MS4A1-high cells")
```

**Expected biology**: Genes elevated in MS4A1-high cells should include other B cell effectors
(e.g., CD79A/B, HLA-DQ genes, IGHM), while depleted genes should reflect non-B cell lineages
(e.g., CD3D for T cells, FCGR3A for monocytes) — consistent with co-expression of the B cell
transcriptional programme rather than random correlation.

### Enrichment

```{r enrichment}
if (length(res$enrichments) == 0 || is.null(res$enrichments)) {
  cat("No significant enrichment returned (may indicate insufficient DEGs).\n")
} else {
  cat("Enrichment result slots:", paste(names(res$enrichments), collapse = ", "), "\n\n")
  # Display the first available enrichment result (top 10 terms)
  first_slot <- res$enrichments[[1]]
  if (!is.null(first_slot) && nrow(first_slot) > 0) {
    top_terms <- head(first_slot[order(first_slot$p.adjust), ], 10)
    knitr::kable(
      top_terms[, intersect(
        c("Description", "GeneRatio", "p.adjust", "Count"),
        colnames(top_terms)
      )],
      digits = 4,
      caption = "Top 10 enriched GO terms in MS4A1-high vs MS4A1-low contrast"
    )
  }
}
```

## Interpretation

If enrichment results are available above, the top GO terms should reflect
biology consistent with MS4A1 (CD20) — a canonical B cell surface marker and
the pharmacological target of Rituximab. Expected functional themes include
adaptive immune response, B cell activation, immunoglobulin production, and
antigen receptor-mediated signalling.

This alignment between FIP output and well-established B cell biology serves as
a **methodological positive control**, demonstrating that the pipeline can
faithfully capture gene-associated transcriptional programmes from endogenous
expression variation — a prerequisite for applying FIP to lesser-characterised
GWAS-prioritised candidates.

> **Caution**: FIP contrasts endogenous expression strata, not experimental perturbations.
> Enriched terms reflect co-expressed programmes, which may partly arise from cell-type
> composition differences. The cell-type composition table above and the inclusion of a
> `filter_cell_threshold` parameter are designed to mitigate — but not eliminate — this confound.
