---
title: "In-silico Knockout (SKO) Demo with Seurat Built-in Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{In-silico Knockout (SKO) Demo with Seurat Built-in Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
leo_ns <- NULL
if (
  requireNamespace("leo.sc", quietly = TRUE) &&
  exists("silico_ko", envir = asNamespace("leo.sc"), inherits = FALSE)
) {
  library(leo.sc)
  leo_ns <- asNamespace("leo.sc")
} else {
  if (!requireNamespace("devtools", quietly = TRUE)) {
    stop("Please install leo.sc or devtools before rendering this tutorial.")
  }
  leo_ns <- devtools::load_all(".", quiet = TRUE)$env
}
silico_ko <- get("silico_ko", envir = leo_ns)
```

## Overview

This tutorial demonstrates a complete SKO run using `silico_ko()` from `leo.sc`.

- Dataset: `SeuratObject::pbmc_small` (built-in Seurat demo object)
- Goal: run a reproducible in-silico knock-out contrast for one immune gene
- Output: matched high/low cell groups, differential expression results, and enrichment results

## Method (text + flowchart)

`silico_ko()` in this demo follows a fixed six-step pipeline:

```text
Expressing-cell filter
  -> Cell-type filter
  -> High-expression group quota select
  -> Matched low-expression group select
  -> DEG (FindMarkers)
  -> Enrichment (leo_enrich)
```

## Data & Gene Setup

```{r data-setup}
data("pbmc_small", package = "SeuratObject")
srt <- pbmc_small

# silico_ko() expects a cell annotation column
srt$cell_anno <- as.character(Seurat::Idents(srt))

candidate_genes <- c("LST1", "NKG7", "TYROBP", "IL7R", "CD3D", "MS4A1")

expr_mat <- tryCatch(
  Seurat::GetAssayData(srt, layer = "data"),
  error = function(e) Seurat::GetAssayData(srt, slot = "data")
)

available_genes <- intersect(candidate_genes, rownames(expr_mat))
stopifnot(length(available_genes) > 0)

gene_positive_counts <- vapply(
  available_genes,
  function(g) sum(expr_mat[g, ] > 0),
  numeric(1)
)

gene_rank <- data.frame(
  gene = names(gene_positive_counts),
  n_positive_cells = as.integer(gene_positive_counts),
  stringsAsFactors = FALSE
)
gene_rank <- gene_rank[order(gene_rank$n_positive_cells, decreasing = TRUE), ]

knitr::kable(gene_rank, caption = "Candidate immune genes ranked by positive-cell count")

gene_use <- gene_rank$gene[[1]]
gene_use
```

## Run SKO

```{r run-sko}
res <- silico_ko(
  all = srt,
  gene = gene_use,
  sko_mode = "ko",
  cell_col = "cell_anno",
  filter_cell_threshold = 3,
  pct_threshold = 0.2,
  deg_method = "default",
  enrichment_method = "ORA",
  enrichment_bg = "GO"
)

names(res)
```

## Interpret Outputs

```{r inspect-output}
length(res$high_cells)
length(res$low_cells)

head(res$cell_rato)

deg <- res$deg_results
deg <- tibble::rownames_to_column(deg, "gene")
head(deg[, c("gene", "avg_log2FC", "p_val_adj")])

if (length(res$enrichments) == 0) {
  cat("No enrichment result was returned (expected for very small demo data).\n")
} else {
  print(names(res$enrichments))
}
```

## Caveats

- `pbmc_small` is intentionally tiny and is only used here to keep the demo lightweight.
- Enrichment output can be empty (`res$enrichments` may be `list()`) and this is expected in small datasets.
- For biological interpretation, use larger datasets and review parameters such as `pct_threshold`, `filter_cell_threshold`, and the enrichment background.
