---
title: "FIP Case Study: TYROBP in PBMC3k"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{FIP Case Study: TYROBP in PBMC3k}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)

find_leo_sc_root <- function(start = getwd()) {
  cur <- normalizePath(start, winslash = "/", mustWork = TRUE)
  repeat {
    desc <- file.path(cur, "DESCRIPTION")
    if (file.exists(desc)) {
      desc_lines <- readLines(desc, warn = FALSE)
      if (any(grepl("^Package:\\s*leo\\.sc\\s*$", desc_lines))) {
        return(cur)
      }
    }
    parent <- dirname(cur)
    if (identical(parent, cur)) {
      return(NULL)
    }
    cur <- parent
  }
}

pkg_root <- find_leo_sc_root()
if (!is.null(pkg_root) && requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else if (!requireNamespace("leo.sc", quietly = TRUE)) {
  stop("Please install leo.sc, or render from a leo.sc source checkout.")
}

silico_ko <- get("silico_ko", envir = asNamespace("leo.sc"), inherits = FALSE)
```

## Context

This case study applies FIP (`silico_ko()`) to the classic **PBMC3k** dataset
([10x Genomics, 2016](https://support.10xgenomics.com/single-cell-gene-expression/datasets/1.1.0/pbmc3k))
to infer transcriptional programs associated with high endogenous expression of **TYROBP** —
a myeloid DAP12 adaptor protein critical for monocyte/macrophage activation and innate immune signalling.

Compared with the [minimal pbmc_small demo](tutorial-sko.html), this vignette uses a
properly annotated 2,700-cell dataset to demonstrate that FIP produces statistically
meaningful DEG and enrichment results.

## Data

```{r data-setup}
# Requires SeuratData >= 0.2.2 (github::satijalab/seurat-data)
if (!requireNamespace("SeuratData", quietly = TRUE)) {
  stop("SeuratData not found. Install with: devtools::install_github('satijalab/seurat-data')")
}

suppressMessages(SeuratData::InstallData("pbmc3k"))
suppressMessages(data("pbmc3k.final", package = "SeuratData"))

# pbmc3k.final has pre-computed clusters and canonical cell-type labels in `seurat_annotations`
# Reference: https://satijalab.org/seurat/articles/pbmc3k_tutorial
srt <- pbmc3k.final
srt$cell_anno <- as.character(srt$seurat_annotations)

cli::cli_alert_info("Dataset: {ncol(srt)} cells × {nrow(srt)} genes | {length(unique(srt$cell_anno))} cell types")

# Overview of cell-type composition
ct_tbl <- as.data.frame(sort(table(srt$cell_anno), decreasing = TRUE))
colnames(ct_tbl) <- c("Cell type", "N cells")
knitr::kable(ct_tbl, caption = "PBMC3k cell-type composition")
```

## FIP: TYROBP Knock-High Analysis

TYROBP (DAP12) is abundantly expressed in **CD14+ Monocytes** and acts as a signalling
scaffold for multiple innate immune receptors. FIP contrasts the high- vs low-expression
groups to reveal associated transcriptional programmes.

```{r run-sko}
res <- silico_ko(
  all = srt,
  gene = "TYROBP",
  sko_mode = "ko", # "ko" = compare high vs low expressors
  cell_col = "cell_anno",
  filter_cell_threshold = 10, # exclude cell types with < 10 expressing cells
  pct_threshold = 0.1, # top/bottom 10% of expressors per cell type
  deg_method = "default", # Wilcoxon rank-sum via Seurat::FindMarkers
  enrichment_method = "ORA", # over-representation analysis
  enrichment_bg = "GO", # GO background
  simplify = TRUE # simplify redundant GO terms
)
```

## Results

### Cell-type composition of g^high group

```{r cell-composition}
knitr::kable(res$cell_rato,
  caption = "Cell-type composition of TYROBP-high group (g^high)"
)
```

*Expected result:* CD14+ Monocytes should dominate the g^high fraction, consistent with
TYROBP's canonical role in myeloid cells.

### Differential expression

```{r deg-summary}
deg <- tibble::rownames_to_column(res$deg_results, "gene")
n_up <- sum(deg$avg_log2FC > 0 & deg$p_val_adj < 0.05, na.rm = TRUE)
n_down <- sum(deg$avg_log2FC < 0 & deg$p_val_adj < 0.05, na.rm = TRUE)
cli::cli_alert_info("DEG summary — upregulated in g^high: {n_up} | downregulated: {n_down}")

top_up <- head(deg[deg$avg_log2FC > 0 & !is.na(deg$p_val_adj), ], 5)
top_down <- head(deg[deg$avg_log2FC < 0 & !is.na(deg$p_val_adj), ], 5)

knitr::kable(top_up[, c("gene", "avg_log2FC", "p_val_adj")],
  digits = 3, caption = "Top genes upregulated in TYROBP-high cells"
)
knitr::kable(top_down[, c("gene", "avg_log2FC", "p_val_adj")],
  digits = 3, caption = "Top genes downregulated in TYROBP-high cells"
)
```

### Enrichment

```{r enrichment}
if (length(res$enrichments) == 0 || is.null(res$enrichments)) {
  cat("No significant enrichment returned (may indicate insufficient DEGs).\n")
} else {
  cat("Enrichment result slots:", paste(names(res$enrichments), collapse = ", "), "\n")
}
```

## Interpretation

FIP outputs a data-driven functional portrait of TYROBP-associated programmes.
Up-regulated transcripts in g^high cells reflect biological processes that co-vary
with high DAP12 expression — including innate immune activation, myeloid differentiation,
and potentially phagocytosis — providing candidate mechanistic hypotheses for
downstream wet-lab validation.

> **Note**: FIP contrasts transcriptional states defined by endogenous expression level.
> Results reflect correlates, not definitive causal effects of TYROBP perturbation.
> See the [FIP method overview](tutorial-sko.html#method) for algorithmic details.
